Appendix S1. Linear regression models and non-linear population dynamics

### 2024-09-11 manuscript/Appendix_S1_v01.Rmd
spatial not working

>>>>

# Species growth, competition, environmental change, and trait evolution in a spatially structured environment

A final consideration to add to the model is the existence of multiple sites that may have correlation structure in the variance of species abundances. We consider this in our simulation by adding multiple (10) sites, with no vs. high spatial covariance. We then consider spatial random effects in HMSC, where covariance is considered as a function of distance between sites.

We use the same example as above, but now with 10 sites. To illustrate how HMSC can estimate the impacts of distance-dependent spatial covariance, we duplicate the population time series obtained in the example above, but we add in site variance and among-site covariance.

# model function
disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
  What <- Wmax*sqrt(w/(P+w))
  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
  return(c(Nt1,Nt2,r1,r2))
}
# Simulation of model for t time steps, i sites
j <- 10
# random site locations
xycoords = matrix(runif(2*j), ncol = 2)
rownames(xycoords) <- 1:j
t <- 40
N <- array(NA,dim=c(j*t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
x <- array(NA,dim=c(j*t,2))
x <- as.data.frame(x)
colnames(x) <- c("x1","x2")
r <- array(NA,dim=c(j*t,2))
r <- as.data.frame(r)
colnames(r) <- c("r1","r2")
N$N1[1:j] <- N1.0
N$N2[1:j] <- N2.0
x$x1[1:j] <- x1.0
x$x2[1:j] <- x2.0
What <- Wmax*sqrt(w/(P+w))
r$r1[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
r$r2[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
E <- rep(NA, t)
E[1] <- E.0
# Keep track of site and time
study <- array(NA,dim=c(j*t,3))
study <- as.data.frame(study)
colnames(study) <- c("obs","site","time")
study$obs <- 1:dim(study)[1]
study$site <- rep(1:j,times=t)
study$time <- rep(1:t,each=j)

for (i in 2:t) {
  for(z in 1:j){
    res <- disc_LV_evol(N1.0=N[study$site==z & study$time==i-1,1],N2.0=N[study$site==z & study$time==i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[study$site==z & study$time==i-1,1],x2=x[study$site==z & study$time==i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
    N$N1[study$site==z & study$time==i] <- res[1]
    N$N2[study$site==z & study$time==i] <- res[2]
    r$r1[study$site==z & study$time==i] <- res[3]
    r$r2[study$site==z & study$time==i] <- res[4]
    # trait change
    d1 <- E[i-1] - x[study$site==z & study$time==i-1,1]
    d2 <- E[i-1] - x[study$site==z & study$time==i-1,2]
    d1.1 <- k*d1
    d2.1 <- k*d2
    x$x1[study$site==z & study$time==i] <- E[i-1] - d1.1
    x$x2[study$site==z & study$time==i] <- E[i-1] - d2.1
  }
  # site-level random effect
  sigma <- 0
  sigma.spatial <- 2
  alpha.spatial <- 0.5
  Sigma = sigma.spatial^2*exp(-as.matrix(dist(xycoords))/alpha.spatial)
  # draw from covariance matrix
  a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
  N$N1[study$time==i] <- N$N1[study$time==i] + a
  a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
  N$N2[study$time==i] <- N$N2[study$time==i] + a
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
}
```
```{r}
# Plot simulation: ggplot
N$site <- study$site
N$time <- study$time
dat <- melt(N, id.vars=c("site","time"))
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2)
```
We now fit a spatial random effect, and add that to the HMSC model.

```{r}
# Plot simulation: ggplot
#N[N <0] <- 0
dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
colnames(dat) <- c("N1", "N2", "x1", "x2")
dat$time <- 1:t
df <- data.frame(cbind(dat$N1[dat$time !=1], dat$N2[dat$time !=1], dat$x1[dat$time !=1], dat$x2[dat$time !=1]))
colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
df$dx1 <- abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t])
df$dx2 <- abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])
# prepare data in HMSC format
Y <- as.matrix(cbind(df$Nt1, df$Nt2))
XData <- data.frame(cbind(rep(E[1:(t-1)],times=j),rep(E[1:(t-1)],times=j)^2,abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t]),abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])))
colnames(XData) <- c("E","Esq","dx1","dx2")
# Update study design to include spatial random effects
studyDesign = data.frame(sample = as.factor(1:(j*(t-1))),time=rep(1:(t-1),each=j),site=rep(1:j,times=(t-1)))
studyDesign$time <- as.factor(studyDesign$time)
studyDesign$site <- as.factor(studyDesign$site)
rL.time = HmscRandomLevel(units = studyDesign$time)
rL.spatial = HmscRandomLevel(sData = xycoords)
# Fit HMSC model
m.9.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list("time" = rL.time,"site"=rL.spatial))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.9.sample <- sampleMcmc(m.9.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

```{r}
m9.post.hmsc <- convertToCodaObject(m.9.sample)
summary(m9.post.hmsc$Beta)
bayesplot::mcmc_trace(m9.post.hmsc$Beta)
bayesplot::mcmc_areas(m9.post.hmsc$Beta,area_method = c("equal height"))
```
```{r warning=FALSE}
VP <- computeVariancePartitioning(m.9.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
plotVariancePartitioning(m.9.sample,VP,cols=c("white","skyblue","darkgrey","orange","black"),args.legend=list(cex=0.75,bg="transparent"))
```
